---
title: "001_diversity_27_thalianas"
output: html_document
date: "2024-03-22"
---

# Upload packages 
```{r setup, include=FALSE}
library(plyr,quietly = T)
library(stringi,quietly = T)
library(data.table,quietly = T)
library(tools,quietly = T)
library(data.table,quietly = T)
library(stringdist,quietly = T)
library(stringr,quietly = T)
library(ggplot2, quietly = T)
library(dplyr, quietly = T)
library(gridExtra,quietly = T)
library(ggpubr,quietly = T)
library(cowplot,quietly = T)
library(patchwork,quietly = T)
library(grid,quietly = T)
library(Cairo,quietly = T)
library(magrittr,quietly = T)
library(tidyr)
library(broom)
library(Cairo)
library(data.table)
library(dplyr)
library(scales)
library(tidyr)
```
# Read data (diversity within A.thaliana)
```{r}
# Read data 
pi_4fold_thal <- fread("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_27g/seq_all_snps.biallelic_and_non_variable.changed_27g_thaliana_4_fold_sites_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
pi_genes_thal <- fread("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_27g/seq_all_snps.biallelic_and_non_variable.changed_27g_thaliana_all_genes_pi_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
pi_all_contexts  <- fread("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_27g/seq_all_snps.biallelic_and_non_variable.changed_27g_thaliana_all_contexts_pi_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
# Read diversity within A.lyrata  
pi_4fold_lyr_pi <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_thaliana_27g_lyrata_31g/thaliana_lyrata_diversity_divergence_4fold_sites_pi.txt", header=T, sep='\t')
pi_4fold_lyr_pi <-pi_4fold_lyr_pi[pi_4fold_lyr_pi$pop=="LYR",]
# Read divergence
pi_4fold_lyr_dxy <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_thaliana_27g_lyrata_31g/thaliana_lyrata_diversity_divergence_4fold_sites_dxy.txt", header=T, sep='\t')

## Read centromeres 
centromeres <- read.csv2('/groups/nordborg/projects/genome_organization/01_data/tair10/centr_bound.txt',header=T, sep=",")
chromosome_lengths <- data.frame(
  chromosome = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"),
  Golden_path_length = c("30427671", "19698289", "23459830", "18585056", "26975502")
)
# Calculate  500 KB widnows 
calc_window <- function(data) {
   data <- data %>%
     mutate(
       window_start = (window_pos_1 %/% 300000) * 300000 + 1,
       window_end = window_start + 299999
     )
   return(data)  # Explicitly return the modified data
}

# Function to average pi to the windows 
average_pi <- function(data){
  data <-data %>% group_by(chromosome, window_start, window_end) %>% dplyr::summarise(avg_pi_window = sum((count_diffs)) / sum(count_comparisons))
  return(data)
}
# Function to calculate gene density
calc_gene_density <- function(data) {
  data <- data %>%
    group_by(chromosome, window_start, window_end) %>%
    dplyr::summarise(gene_count = n())
  return(data)
}
```
## Calculate windows 
```{r}
# Calculate windows  
pi_4fold_thal <- calc_window(pi_4fold_thal)
pi_genes_thal <- calc_window(pi_genes_thal)
pi_all_contexts <- calc_window(pi_all_contexts)
pi_lyrata <- calc_window(pi_4fold_lyr_pi)
divergence <- calc_window(pi_4fold_lyr_dxy)
# Calculate gene density 
pi_4fold_thal_gene_den <- calc_gene_density(pi_4fold_thal)
pi_genes_thal_gene_den <- calc_gene_density(pi_genes_thal)
pi_all_contexts_gene_den <- calc_gene_density(pi_all_contexts)
pi_lyrata_gene_den <- calc_window(pi_lyrata)
divergence_gene_den <- calc_window(divergence)
# Average pi 
pi_4fold_thal <- average_pi(pi_4fold_thal)
pi_genes_thal <- average_pi(pi_genes_thal)
pi_all_contexts <- average_pi(pi_all_contexts)
pi_lyrata <- average_pi(pi_lyrata)
divergence <- average_pi(divergence)
```
## Combine dataframes 
```{r}
# Combine dataframes based on the windows 
pi_4fold_thal <- pi_4fold_thal %>% dplyr::select(chromosome,window_start,window_end,avg_pi_window)
pi_genes_thal <- pi_genes_thal %>% dplyr::select(chromosome,window_start,window_end,avg_pi_window)
pi_all_contexts <- pi_all_contexts %>% dplyr::select(chromosome,window_start,window_end,avg_pi_window)

# Combine pi into one dataframe
pi_combined <- pi_all_contexts %>%
  full_join(pi_4fold_thal, by = c("chromosome", "window_start", "window_end"), suffix=c("_all_cont","_4_fold")) %>%
  full_join(pi_genes_thal %>% rename_with(~paste0(.x, "_genes"), -c(chromosome, window_start, window_end)), 
            by = c("chromosome", "window_start", "window_end")) %>%
  full_join(pi_lyrata %>% rename_with(~paste0(.x, "_lyr"), -c(chromosome, window_start, window_end)), 
            by = c("chromosome", "window_start", "window_end")) %>%
  full_join(divergence %>% rename_with(~paste0(.x, "_divergence"), -c(chromosome, window_start, window_end)), 
            by = c("chromosome", "window_start", "window_end")) 
```

# Add gene density 
```{r}
pi_4fold_thal_gene_den <- pi_4fold_thal_gene_den %>% dplyr::select(chromosome,window_start,window_end,gene_count)
pi_genes_thal_gene_den <- pi_genes_thal_gene_den %>% dplyr::select(chromosome,window_start,window_end,gene_count)
pi_all_contexts_gene_den <- pi_all_contexts_gene_den %>% dplyr::select(chromosome,window_start,window_end,gene_count)
# Merge with gene density 

all_gene_density <- pi_all_contexts_gene_den %>%
  full_join(pi_4fold_thal_gene_den, by = c("chromosome", "window_start", "window_end"), suffix=c("_all_cont","_4_fold")) %>%
  full_join(pi_genes_thal_gene_den, by = c("chromosome", "window_start", "window_end"),suffix=c("_tes","_genes"))

# Merge with pi 
pi_combined_with_gene_density <- inner_join(all_gene_density,pi_combined,by = c("chromosome", "window_start", "window_end"))
```
# Generate expected windows 
```{r}
generate_windows <- function(chromosome, length) {
  tibble(
    chromosome = chromosome,
    window_start = seq(1, length, by = 300000),
    window_end = (window_start + 300000 - 1))}
# Apply this function to each chromosome to create a complete set of expected windows
expected_windows <- chromosome_lengths %>%
  rowwise() %>%
  do(generate_windows(.$chromosome, .$Golden_path_length)) %>%
  ungroup()

# Impute windows which are not in the data (to have 0)
data_full_range <- expected_windows %>%
  full_join(pi_combined, by = c("chromosome", "window_start", "window_end")) 

# Merge with centromere
data_full_range_raw <- merge(data_full_range, centromeres, by="chromosome")
# Make everything in centromere 0
# Adjust pi values if window falls within the centromere range
data_full_range_centr <- data_full_range_raw %>%
  dplyr::mutate(
    avg_pi_window_all_cont = ifelse(window_end >= start & window_start <= stop, "NA", avg_pi_window_all_cont),
    avg_pi_window_4_fold   = ifelse(window_end >= start & window_start <= stop, "NA", avg_pi_window_4_fold),
    avg_pi_window_genes    = ifelse(window_end >= start & window_start <= stop,"NA", avg_pi_window_genes),
    avg_pi_window_lyr       = ifelse(window_end >= start & window_start <= stop, "NA", avg_pi_window_lyr),
    avg_pi_window_divergence   = ifelse(window_end >= start & window_start <= stop, "NA", avg_pi_window_divergence)
  )
```

```{r}
# All colorblind-friendly
display_carto_all(colorblind_friendly = TRUE)

# Specific type
display_carto_all(type = "qualitative", colorblind_friendly = TRUE)
display_carto_all(type = "diverging", colorblind_friendly = TRUE)

carto_pal(n = 7, name = "Safe")
```

## Create vertical facets 
```{r}
median_all     <- median(data_full_range_centr$avg_pi_window_all_cont, na.rm = TRUE)
median_genes   <- median(data_full_range_centr$avg_pi_window_genes,     na.rm = TRUE)
median_4_fold  <- median(data_full_range_centr$avg_pi_window_4_fold,    na.rm = TRUE)
median_lyrata  <- median(data_full_range_centr$avg_pi_window_lyr,       na.rm = TRUE)
median_dxy     <- median(data_full_range_centr$avg_pi_window_divergence,na.rm = TRUE)

data_long <- data_full_range_centr %>%
  select(chromosome, window_start, window_end, 
         avg_pi_window_4_fold, avg_pi_window_lyr, avg_pi_window_divergence,
         start, stop) %>%
  pivot_longer(cols = c(avg_pi_window_4_fold, avg_pi_window_lyr, avg_pi_window_divergence),
               names_to = "measure",
               values_to = "value")

# Create factor for measure to control order
data_long$measure <- factor(data_long$measure,
                            levels = c("avg_pi_window_4_fold", 
                                      "avg_pi_window_lyr", 
                                      "avg_pi_window_divergence"),
                            labels = c("π (4-fold)", "A. lyrata π (4-fold)", "Dxy (4-fold)"))

# Create centromere data frame (unique per chromosome)
centromere_data <- data_full_range_raw %>%
  select(chromosome, start, stop) %>%
  distinct()

# Create median data frame
median_data <- data.frame(
  measure = factor(c("π (4-fold)", "A. lyrata π (4-fold)", "Dxy (4-fold)"),
                   levels = c("π (4-fold)", "A. lyrata π (4-fold)", "Dxy (4-fold)")),
  median_value = c(median_4_fold, median_lyrata, median_dxy)
)

# Create the plot with colors
p <- ggplot(data_long, aes(x = value, y = window_start/1e6, color = measure, group = measure)) +
  geom_vline(data = median_data, aes(xintercept = median_value), 
             linetype = "dashed", color = "black", size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, size = 0.6, alpha = 0.6, span = 0.1, fill = "gray62", orientation = "y") +
  geom_point(size = 0.8, alpha = 0.6) + 
  facet_grid(chromosome ~ measure, scales = "free") +
  scale_color_manual(values = c(
    "π (4-fold)" = "#117733",
    "A. lyrata π (4-fold)" = "#332288",
    "Dxy (4-fold)" = "#AA4499"
  )) + geom_rect(data = centromere_data, 
            aes(xmin = -Inf, xmax = Inf, ymin = start/1e6, ymax = stop/1e6),
            fill = "gray62", alpha = 0.5, inherit.aes = FALSE) +
  theme_void() +
  theme(strip.text.x = element_text(size = 20, margin = margin(t = 10, b = 10)),
        strip.text.y = element_text(size = 20, angle = 270, margin = margin(l = 10, r = 10)),
        axis.text.x = element_text(size = 16,angle = 45, hjust = 1),
        axis.text.y = element_text(size = 16,angle = 90, margin = margin(r = 5)),
        axis.title.x = element_text(size = 20, margin = margin(t = 10)),
        axis.title.y = element_text(size = 20, angle = 90, margin = margin(r = 15)),
        axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y.left = element_line(color = "black", size = 0.5),
        panel.spacing = unit(0.5, "lines"),
        legend.position = "none") +
  labs(x = "Value", y = "Position (Mbp)")

width_inches <- 12
height_inches <- 15
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/chromosome_plots_colored.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(p)
dev.off()
```
## Make scatterplots 
```{r}

library(ggplot2)
library(dplyr)

# Calculate distance to centromere
data_full_range_raw <- data_full_range_raw %>%
  mutate(
    window_mid = (window_start + window_end) / 2,
    centromere_mid = (start + stop) / 2,
    dist_to_centromere = abs(window_mid - centromere_mid),
    chr_num = as.numeric(gsub("Chr", "", chromosome))
  ) %>%
  arrange(chr_num, window_start)

# Plot
ggplot(data_full_range_raw, aes(x = avg_pi_window_4_fold, y = avg_pi_window_divergence)) +
  geom_point(aes(color = dist_to_centromere/1e6), alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "gray62", linetype = "dashed") +
  scale_color_viridis_c(option = "plasma", direction = -1,
                        name = "Distance to\ncentromere (Mbp)") +
  labs(x = "A. thaliana π (4-fold)", 
       y = "Divergence (4-fold)") +
  theme_void() + 
  theme(
    axis.text.x = element_text(size = 25, angle = 45, hjust = 1, margin = margin(t = 10)),
    axis.text.y = element_text(size = 25, margin = margin(r = 10)),
    strip.text = element_text(size = 25, margin = margin(b = 10)),
    axis.title.x = element_text(size = 25, margin = margin(t = 20)),
    axis.title.y = element_text(size = 25, angle = 90, margin = margin(r = 20)),
    axis.line = element_line(color = "black"),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.key.height = unit(1.5, "cm"),
    legend.key.width = unit(0.8, "cm"),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) + 
#print slope coefficient 
  #facet_wrap(~chromosome, ncol = 5, scales = "free")
```

### Dotplot for each chromosome
```{r}
chr1_pos <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/001_pannagram_31_lyr_thal/out/rearragned_genomes/out/features/msa/extracted_MN47_TAIR10_chr5.tsv", header=T)
```

```{r}
bin_size <- 10000  # Adjust based on your genome size
library(dplyr)

chr1_pos_filt <- chr1_pos %>%
  filter(GCA_944990045.1_MN47_v2_genomic.8scaff != 0 & TAIR10 > 0)

chr1_binned <- chr1_pos_filt %>%
  mutate(
    TAIR10_bin = floor(TAIR10 / bin_size) * bin_size,
    MN47_bin = floor(abs(GCA_944990045.1_MN47_v2_genomic.8scaff) / bin_size) * bin_size
  ) %>%
  group_by(TAIR10_bin, MN47_bin) %>%
  summarise(count = n(), .groups = "drop")

# Plot
ggplot(chr1_binned, aes(x = MN47_bin/1e6, y = TAIR10_bin/1e6)) +
  geom_point(size = 2, alpha = 0.8, color = "#352F44") +
  labs(x = "TAIR10 Position (Mbp)", 
       y = "MN47 Position (Mbp)") +
  theme_void() +   theme(
    axis.text.x = element_text(size = 25, angle = 45, hjust = 1, margin = margin(t = 10)),
    axis.text.y = element_text(size = 25, margin = margin(r = 10)),
    strip.text = element_text(size = 25, margin = margin(b = 10)),
    axis.title.x = element_text(size = 25, margin = margin(t = 20)),
    axis.title.y = element_text(size = 25, angle = 90, margin = margin(r = 20)),
  )+
  theme(panel.grid = element_blank(),
        axis.line = element_line(color = "black"))

```

###OLD
### Create in separate facets 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
## Calculate median 
median_all     <- median(data_full_range$avg_pi_window_all_cont, na.rm = TRUE)
median_genes   <- median(data_full_range$avg_pi_window_genes,     na.rm = TRUE)
median_4_fold  <- median(data_full_range$avg_pi_window_4_fold,    na.rm = TRUE)
median_lyrata  <- median(data_full_range$avg_pi_window_lyr,       na.rm = TRUE)
median_dxy     <- median(data_full_range$avg_pi_window_divergence,na.rm = TRUE)
# Reshape to long format
pi_long <- data_full_range %>%
  pivot_longer(
    cols = c(avg_pi_window_all_cont,
             avg_pi_window_4_fold,
             avg_pi_window_genes,
             avg_pi_window_lyr,
             avg_pi_window_divergence),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = case_when(
      metric == "avg_pi_window_all_cont"       ~ "π (genes+TEs)",
      metric == "avg_pi_window_genes"          ~ "π (genes)",
      metric == "avg_pi_window_4_fold"         ~ "π (4 fold sites)",
      metric == "avg_pi_window_lyr"            ~ "A. lyrata π (4 fold sites)",
      metric == "avg_pi_window_divergence"     ~ "Dxy (4 fold sites)"
    ),
    median_value = case_when(
      metric == "π (genes+TEs)"        ~ median_all,
      metric == "π (genes)"            ~ median_genes,
      metric == "π (4 fold sites)"     ~ median_4_fold,
      metric == "A. lyrata π (4 fold sites)"       ~ median_lyrata,
      metric == "Dxy (4 fold sites)"  ~ median_dxy
    )
  )
```

```{r}
# Plot: y = value (numeric), not y = metric

# Set the order you want
pi_long <- pi_long %>%
  mutate(
    metric = factor(
      metric,
      levels = c(
        "π (genes+TEs)",
        "π (genes)",
        "π (4 fold sites)",
        "A. lyrata π (4 fold sites)",
        "Dxy (4 fold sites)"
      )
    )
  )

pi_long_chr1 <- pi_long[pi_long$chromosome=="Chr1",]
plot1 <- ggplot(pi_long_chr1, aes(window_start, value, color = metric)) +
  geom_smooth(method = "loess", se = TRUE, size = 1, alpha = 0.6, span = 0.3, fill = "#C9CDCF") +
  geom_point(data = subset(pi_long_chr1, value != 0), size = 1.3, alpha = 0.9) +
  geom_hline(aes(yintercept = median_value),
             color = "darkgrey", linetype = "dashed", linewidth = 0.6, alpha = 0.9) +
  facet_wrap(metric ~ chromosome, scales = "free") +
  scale_color_manual(values = c(
    "π (genes+TEs)" = "#6A1E55",
    "π (genes)"     = "#D1A980",
    "π (4 fold sites)" = "#4A9782",
    "A. lyrata π (4 fold sites)"   = "#5C5470",
    "Dxy (4 fold sites)" = "#211C84"
  )) +
  theme_bw(base_size = 16) +
  theme(panel.spacing.x = unit(0, "lines"),
        panel.spacing.y = unit(0, "lines"),
        legend.position = "none") +
  xlab("Window Start (Mbp)") +
  ylab("Value") +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6))
plot1

width_inches <- 12
height_inches <- 8
#dpi <- 600
#CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/only_diversity_all_context_figure1_patterns_separately.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
#print(plot1)
dev.off()
```
## Add 

```{r}
library(dplyr)
library(tidyr)

#data_full_range_filtered <-data_full_range[data_full_range$avg_pi_window_divergence>0 & data_full_range$avg_pi_window_all_cont>0,]

library(ggplot2)
library(dplyr)

# Calculate correlations for each chromosome
data_full_range_filtered_chr1 <-data_full_range_filtered[data_full_range_filtered$chromosome=="Chr1",]
correlations <- data_full_range_filtered_chr1 %>%
  group_by(chromosome) %>%
  dplyr::summarise(
    correlation = cor(avg_pi_window_divergence, avg_pi_window_all_cont, use = "complete.obs"),
    .groups = 'drop'
  ) %>%
  mutate(
    cor_label = paste0("r = ", round(correlation, 3)),
    # Position labels in upper left of each facet
    x_pos = -Inf,
    y_pos = Inf
  )

# Create the plot
# Create the plot
ggplot(data_full_range_filtered_chr1, aes(x = avg_pi_window_divergence, y = avg_pi_window_all_cont)) +
  geom_point(aes(fill = avg_pi_window_divergence), 
             alpha = 0.7, 
             size = 4, 
             shape = 21,          # This is the ggplot2 equivalent of pch=21
             color = "black") +   # Border color
  scale_fill_viridis_c(name = "Dxy") +  # Use scale_fill since we're using fill aesthetic
  geom_text(data = correlations, 
            aes(x = x_pos, y = y_pos, label = cor_label),
            hjust = -0.1, vjust = 1.2, 
            size = 8, color = "black", fontface = "plain") +
  theme_bw() +
  facet_wrap(~chromosome, ncol = 5) +
  labs(
    x = "Dxy",
    y = "π (genes+TEs)"
  )+theme_bw(base_size = 20)

# To make it even wider, you can also adjust the figure size when saving:
# ggsave("plot.png", width = 15, height = 10, dpi = 300)

```


## Save diversity table 
```{r}
write.table(pi_long,"/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/pi_27g/diversity_genes_all_cont_and_4fold_sites.csv", quote = F, row.names = F, sep="\t")
```








# Plot diversity in all contexts 
```{r}
plot1 <- ggplot(data_full_range_centr, aes(x = window_start)) +
  geom_line(aes(y = avg_pi_window_all_cont), size = 3, alpha = 0.4, colour = "#643A6B") + 
  geom_line(aes(y = avg_pi_window_4_fold), size = 1, alpha = 0.9, colour = "#78A083") + 
  geom_line(aes(y = avg_pi_window_genes), size = 1, alpha = 0.8, colour = "#ED7D31") +
  facet_wrap(~chromosome, ncol = 5, scales = "free_x") +
  geom_rect(
    data = data_full_range_centr,
    aes(
      xmin = start - 120000, xmax = stop + 120000,
      ymin = -Inf, ymax = Inf
    ), fill = "lightgrey", alpha = 0.1, inherit.aes = FALSE
  ) +
  geom_hline(aes(yintercept = median(avg_pi_window_all_cont, na.rm = TRUE)), 
             linetype = "dashed", colour = "black") + 
  theme_classic() +
  theme(
    panel.border = element_blank(), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.line = element_line(colour = "black"),
    strip.text.x = element_text(size = 10, face = "bold"),
    strip.text.y = element_blank(),
    legend.position = "none"
  ) +
  xlab("Window Start (Mbp)")+scale_x_continuous(labels = label_number(scale = 1e-6))


#width_inches <- 20
#height_inches <- 5
#dpi <- 600
#CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/only_diversity_all_context_figure1_edited.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(plot1)
dev.off()
```



```{r}
pi_combined_clean <- pi_combined %>% 
  select(avg_pi_window_all_cont, avg_pi_window_genes, avg_pi_window_4_fold) %>% 
  drop_na()

median_all <- median(pi_combined_clean$avg_pi_window_all_cont, na.rm = TRUE)
median_genes <- median(pi_combined_clean$avg_pi_window_genes, na.rm = TRUE) 
median_4_fold <- median(pi_combined_clean$avg_pi_window_4_fold, na.rm = TRUE)

# Create your long data
pi_long_filtered <- pi_combined %>%
  pivot_longer(
    cols = c(avg_pi_window_all_cont, avg_pi_window_4_fold, avg_pi_window_genes),
    names_to = "metric",
    values_to = "Diversity (π)"
  ) %>%
  mutate(
    metric = case_when(
      metric == "avg_pi_window_all_cont" ~ "All contexts",
      metric == "avg_pi_window_genes" ~ "Genes",
      metric == "avg_pi_window_4_fold" ~ "4-fold sites"
    ),
    median_value = case_when(
      metric == "All contexts" ~ median_all,
      metric == "Genes" ~ median_genes,
      metric == "4-fold sites" ~ median_4_fold
    )
  )

# Plot
pi_long <- pi_long_filtered %>%
  mutate(
    metric = factor(metric, levels = c("All contexts", "Genes", "4-fold sites"))
  )

pi_long_filtered <-pi_long_filtered[pi_long_filtered$`Diversity (π)`<0.03,]
pi_long_filtered <- left_join(pi_long_filtered, centromeres, by="chromosome")
plot1 <- ggplot(pi_long, aes(x = window_start, y = `Diversity (π)`, color = metric)) +
  geom_point(size = 1.3, alpha = 0.9) +
  geom_smooth(method = "loess", se = FALSE, size = 1.4, alpha = 0.8, span = 0.3,fill = "grey50") +
  geom_hline(aes(yintercept = median_value, color = "darkgrey"), 
             linetype = "dashed", size = 0.6, alpha = 0.9)+
  facet_grid(metric ~ chromosome, scales = "free") +
  scale_color_manual(values = c(
    "All contexts" = "#004030",
    "4-fold sites" = "#D1A980", 
    "Genes" = "#4A9782"
  )) +
 # geom_rect(
#    data = data_full_range ,
#    aes(xmin = start - 150000, xmax = stop + 150000, ymin = -Inf, ymax = Inf), 
#    fill = "lightgrey", alpha = 0.3, inherit.aes = FALSE
#  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.spacing.x = unit(0, "lines"), 
    panel.spacing.y = unit(0, "lines"),
    legend.position = "none"
  ) +
  xlab("Window Start (Mbp)") +
  scale_x_continuous(labels = label_number(scale = 1e-6))

width_inches <- 12
height_inches <- 8
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/only_diversity_all_context_figure1_patterns_separately.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(plot1)
dev.off()
```



```{r}
pi_long_filtered <- na.omit(pi_long_filtered)
# Create a dataset that labels each window as left arm, centromere, or right arm
pi_filtered_arms <- pi_long_filtered %>%
  mutate(arm = case_when(
    window_end < start ~ "left",
    window_start > stop ~ "right",
    TRUE ~ "centromere"
  )) %>%
  filter(arm != "centromere")

ggplot(pi_long_filtered, aes(x = window_start/1e6, y = `Diversity (π)`, color = metric)) +
  geom_point(size = 1.3, alpha = 0.9) +
  # Add vertical stripes for centromeres
  geom_rect(
    data = pi_long_filtered %>% distinct(chromosome, start, stop),
    aes(xmin = (start-100000)/1e6, xmax = (stop+200000)/1e6, ymin = -Inf, ymax = Inf), 
    fill = "lightgrey", alpha = 0.3, inherit.aes = FALSE
  ) +
  # Break geom_smooth at centromeres by creating separate smooths for each arm
  geom_smooth(
    data = pi_filtered_arms %>% mutate(window_start_mbp = window_start/1e6),
    aes(x = window_start_mbp, group = interaction(metric, chromosome, arm)),
    method = "loess", se = FALSE, size = 1.4, alpha = 0.8, span = 0.3
  ) +
  geom_hline(aes(yintercept = median_value, color = "darkgrey"), 
             linetype = "dashed", size = 0.6, alpha = 0.9) +
  facet_grid(metric ~ chromosome, scales = "free") +
  scale_color_manual(values = c(
    "All contexts" = "#004030",
    "4-fold sites" = "#D1A980", 
    "Genes" = "#4A9782"
  )) +
  theme_minimal(base_size = 16) +
  theme(
    panel.spacing.x = unit(0, "lines"), 
    panel.spacing.y = unit(0, "lines"),
    legend.position = "none"
  ) +
  xlab("Window Start (Mbp)")




pi_long_filtered %>% filter(chromosome=="Chr1") %>% ggplot(pi_long_filtered, aes(x = window_start/1e6, y = `Diversity (π)`, color = metric, fill = metric)) +
  geom_point(size = 1.3, alpha = 0.9) +
  # Add vertical stripes for centromeres
  geom_rect(
    data = pi_long_filtered %>% distinct(chromosome, start, stop),
    aes(xmin = (start-100000)/1e6, xmax = (stop+200000)/1e6, ymin = -Inf, ymax = Inf), 
    fill = "lightgrey", alpha = 0.3, inherit.aes = FALSE
  ) +
  # Add ribbon for area above median
  geom_ribbon(
    data = pi_filtered_arms %>% mutate(window_start_mbp = window_start/1e6),
    aes(x = window_start_mbp, 
        ymin = median_value, 
        ymax = pmax(`Diversity (π)`, median_value),
        group = interaction(metric, chromosome, arm)),
    alpha = 0.2, color = NA
  ) +
  # Break geom_smooth at centromeres by creating separate smooths for each arm
  geom_smooth(
    data = pi_filtered_arms %>% mutate(window_start_mbp = window_start/1e6),
    aes(x = window_start_mbp, group = interaction(metric, chromosome, arm)),
    method = "loess", se = FALSE, size = 1.4, alpha = 0.8, span = 0.3
  ) +
  geom_hline(aes(yintercept = median_value, color = "darkgrey"), 
             linetype = "dashed", size = 0.6, alpha = 0.9) +
  facet_grid(metric ~ chromosome, scales = "free") +
  scale_color_manual(values = c(
    "All contexts" = "#004030",
    "4-fold sites" = "#D1A980", 
    "Genes" = "#4A9782"
  )) +
  scale_fill_manual(values = c(
    "All contexts" = "#004030",
    "4-fold sites" = "#D1A980", 
    "Genes" = "#4A9782"
  )) +
  theme_minimal(base_size = 14) +
  theme(
    panel.spacing.x = unit(0, "lines"), 
    panel.spacing.y = unit(0, "lines"),
    legend.position = "none"
  ) +
  xlab("Window Start (Mbp)")
```


```{r}
pi_long_filtered <- pi_long %>%
  ungroup() %>%
  left_join(data_full_range, by = "chromosome")


pi_long_filtered <- pi_long_filtered %>%
  mutate(`Diversity (π)` = ifelse(window_start.x >= (start - 1000) & 
                                  window_start.x <= (stop + 1000), 
                                  0, `Diversity (π)`)) %>%
  select(-contains(".y"), -start, -stop, -gene_count_all_cont, -gene_count_4_fold, 
         -gene_count, -avg_pi_window_all_cont, -avg_pi_window_4_fold, -avg_pi_window,
         -ancestral_start, -ancestral_stop) %>%
  rename(window_start = window_start.x, window_end = window_end.x)
```


## Add divergence

```{r}
pi_combined_with_divergence <- left_join(pi_long_filtered,divergence,by=c("chromosome","window_start", "window_end"))
```
## Plot diversity and divergence
```{r}
# Compute scaling factor
scale_factor <- mean((na.omit(pi_combined_with_divergence$avg_pi_window_4_fold)) / (na.omit(pi_combined_with_divergence$avg_pi_window)))

ggplot(pi_combined_with_divergence, aes(x = window_start)) +
  # Primary Y-axis line (4-fold sites)
  geom_line(aes(y = avg_pi_window_4_fold, colour = "4-fold ortholgous genes diversity"), size = 1, alpha = 0.6) + 
  # Secondary Y-axis line (All Sites, scaled)
  geom_line(aes(y = avg_pi_window/11, colour = "Divergence"), size = 1, alpha = 0.9) + 
  facet_wrap(~chromosome, ncol = 5, scales = "free_x") +
  
  # Add gray highlight regions
  geom_rect(
    data = long_data,
    aes(
      xmin = start - 120000, xmax = stop + 120000,
      ymin = -Inf, ymax = Inf
    ), fill = "lightgrey", alpha = 0.1, inherit.aes = FALSE
  ) +
  
  # Add median line
  geom_hline(aes(yintercept = median(avg_pi_window_4_fold, na.rm = TRUE)), 
             linetype = "dashed", colour = "black") + 
  
  # Define primary and secondary y-axes
  scale_y_continuous(
    name = "Avg π (4-fold sites)", 
    sec.axis = sec_axis(~ . / scale_factor, name = "Avg π (All Sites)")
  ) +
  
  # Set colors for the lines
  scale_colour_manual(
    name = "Legend",
    values = c("Divergence" = "#643A6B", "4-fold ortholgous genes diversity" = "#78A083")
  ) +
  
  # Theme adjustments
  theme_classic() +
  theme(
    panel.border = element_blank(), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.line = element_line(colour = "black"),
    strip.text.x = element_text(size = 10, face = "bold"),
    strip.text.y = element_blank(),
    legend.position = "top"
  ) +
  
  # X-axis formatting
  xlab("Window Start (Mbp)") +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6))





```
## Scatterplot diversity divergence 
```{r}
# Create dataset with centromere proximity
pi_scatter <- pi_combined_with_divergence %>%
  filter(metric == "4-fold sites") %>%
  mutate(
    # Calculate distance to centromere (negative = before, positive = after)
    dist_to_centromere = case_when(
      window_end < start ~ start - window_end,  # left arm
      window_start > stop ~ window_start - stop,  # right arm
      TRUE ~ 0  # inside centromere
    ),
    # Use absolute distance for coloring
    abs_dist_to_centromere = abs(dist_to_centromere)
  )

ggplot(pi_scatter, aes(x = `Diversity (π)`, y = avg_pi_window_divergence, 
                       color = abs_dist_to_centromere)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_c(
    option = "plasma",
    direction = -1,
    name = "Distance to\ncentromere (bp)",
    labels = scales::comma
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right"
  ) +
  xlab("Diversity (π) - 4-fold sites") +
  ylab("Divergence") +
  labs(title = "Diversity vs Divergence by Centromere Proximity")
```

# Read Divergence and recombination rate 
```{r}
pi_4fold_lyr_dxy <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/dxy_27_thal_3lyr/27thal_3lyr_pacbio_dxy.txt", header=T, sep='\t')
pi_4fold_lyr_dxy <-pi_4fold_lyr_dxy[pi_4fold_lyr_dxy$pop1=="POP1" & pi_4fold_lyr_dxy$pop2=="LYR",]
pi_4fold_lyr_dxy <- calc_window(pi_4fold_lyr_dxy)

pi_4fold_lyr_dxy <- pi_4fold_lyr_dxy %>% group_by(chromosome, window_start, window_end) %>% dplyr::summarise(avg_dxy_window = sum((count_diffs)) / sum(count_comparisons))

historical_recombination_chr1 <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr1_ldhat_output_with_loci_adjusted_windows_500kb.csv", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
historical_recombination_chr2 <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr2_ldhat_output_with_loci_adjusted_windows_500kb.csv", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
historical_recombination_chr3 <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr3_ldhat_output_with_loci_adjusted_windows_500kb.csv", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
historical_recombination_chr4 <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr4_ldhat_output_with_loci_adjusted_windows_500kb.csv", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
historical_recombination_chr5 <- read.table("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr5_ldhat_output_with_loci_adjusted_windows_500kb.csv", header=TRUE, sep='\t') %>% na.omit() %>% distinct()

historical_recombination <- rbind(historical_recombination_chr1,historical_recombination_chr2,historical_recombination_chr3,historical_recombination_chr4,historical_recombination_chr5)
```
## Join historical recombination, dxy and pi 
```{r}
historical_recombination_dxy <- inner_join(pi_4fold_lyr_dxy,historical_recombination, by=c("chromosome"="chromosome","window_start"="Start", "window_end"="End"))
# Join with pi in all context 
historical_recombination_dxy_pi <- inner_join(data_full_range,historical_recombination_dxy, by=c("chromosome","window_start","window_end"))
```
# Impute 0 on the places of centromere to the dxy and rho 
```{r}
# Make everything in centromere 0
# Adjust pi values if window falls within the centromere range
historical_recombination_dxy_pi <- historical_recombination_dxy_pi %>% dplyr::mutate(avg_dxy_window = ifelse(window_end >= start & window_start <= stop, 0, avg_dxy_window))
historical_recombination_dxy_pi <- historical_recombination_dxy_pi %>% dplyr::mutate(Mean_rho_mean = ifelse(window_end >= start & window_start <= stop, 0, Mean_rho_mean))
```

```{r}

long_data <- historical_recombination_dxy_pi %>%
  pivot_longer(
    cols = c(avg_pi_window_all_cont, avg_dxy_window, Mean_rho_mean, gene_count_genes),
    names_to = "estimate",
    values_to = "value"
  )

long_data$estimate <- factor(long_data$estimate, levels = c(
  "avg_pi_window_all_cont", "avg_dxy_window", "Mean_rho_mean", "gene_count_genes"
))

# Create the plot
custom_colors <- c(
  "avg_pi_window_all_cont" = "#643A6B",
  "avg_dxy_window" = "#78A083",
  "Mean_rho_mean" = "#ED7D31",
  "gene_count_genes" = "#2B78E4"
)

# Plot with custom colors

plot1 <- ggplot(long_data, aes(x = window_start, y = value, colour = estimate)) +
  geom_rect(
    data = long_data,
    aes(
      xmin = start - 100000, xmax = stop + 100000,
      ymin = -Inf, ymax = Inf, inherit.aes = FALSE # Full height of the plot
    ),
    fill = "lightgrey", alpha = 0.1, inherit.aes = FALSE # Prevent the stripe from inheriting main plot aesthetics
  ) +
  geom_line(size = 1, alpha = 0.8) +
  facet_grid(estimate ~ chromosome, scales = "free_y") +
  scale_color_manual(values = custom_colors) +
  scale_x_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6)) +
  labs(
    x = "Window Start",
    y = "Values"
  ) +
   theme_half_open() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))+
  theme(
    strip.text.x = element_text(size = 10, face = "bold"),
    strip.text.y = element_blank(),
    legend.position = "none")
    #panel.spacing = unit(0, 'lines'))+ theme(panel.spacing.x=unit(1, "lines"))
 

width_inches <- 15
height_inches <- 10
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/diversity_rho_divergence_gene_density_figure1_edited.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(plot1)
dev.off()
   
```




















# old code
```{r}
minimal_theme_settings <- ggplot2::theme_minimal(base_size = 14)
custom_theme <- theme(
  plot.background = element_blank(), 
  strip.text = element_text(size = 14, color = "gray30"),
  panel.background = element_blank(),
  panel.border = element_blank(), # Ensure no frame
  panel.grid = element_blank(),
  axis.title.x = element_text(color = "gray30", size = 14), # Set x-axis title color and size
  axis.title.y = element_text(color = "gray30", size = 14), # Set y-axis title color and size
  axis.ticks.x = element_blank(),
  axis.line = element_line(color = "gray30"), # Add this line to show x and y axis lines
  strip.background = element_blank(),
   axis.text.x = element_text(color = "gray30", size = 14), # Adjust x-axis text size and color
    axis.text.y = element_text(color = "gray30", size = 8),
  strip.text.x = element_text(color = "gray30",size = 14), # Control color of facet wrap titles
  legend.text = element_text(color = "gray30",size = 14) # Set legend text color
)
max_chrom_length <- max(data_full_range$window_start, na.rm = TRUE)
mean_avg_pi_4deg <- mean(data_full_range$avg_pi_4_fold_window, na.rm = TRUE)
#Calculate sd


data_full_range <- data_full_range %>%
  group_by(chromosome) %>%
  mutate(sd = sd(avg_pi_4_fold_window, na.rm = TRUE)) %>%
  ungroup() 

chrom1_data <- data_full_range %>% filter(chromosome == "Chr1")

final_plot <- ggplot(data_full_range, aes(x = window_start, y = avg_pi_4_fold_window)) +
  geom_rect(data = data_full_range, aes(xmin = start-500000 , xmax = stop+500000, ymin = -Inf, ymax = Inf), fill = "#DDDDDD", alpha = 0.03) +
 # geom_rect(aes(xmin = ifelse(start - 1000000 < 1, 1, start ), xmax = stop , ymin = -Inf, ymax = Inf), fill = "#DDDDDD", alpha = 0.03) +
  #geom_vline(data = chrom1_data, aes(xintercept = ancestral_stop), color = "gray30", linetype = "solid", size = 0.8, alpha = 0.3) +
  geom_ribbon(aes(ymin = avg_pi_4_fold_window - sd, ymax = ifelse(avg_pi_4_fold_window > mean_avg_pi_4deg, avg_pi_4_fold_window + sd, avg_pi_4_fold_window - sd)), fill = "#527853", alpha = 0.5) +
  #geom_vline(aes(xintercept = ifelse(start - 1000000 < 1, 1, start )), color = "gray30", linetype = "solid", size = 0.4, alpha = 0.3) +
  #geom_vline(aes(xintercept = stop ), color = "gray30", linetype = "solid", size = 0.8, alpha = 0.3) +
  #geom_point(size = 2, alpha = 0.7, colour = "#527853") +
  geom_line(size = 1, alpha = 0.7, colour = "#527853") +
  geom_hline(aes(yintercept = mean_avg_pi_4deg), color = "#527853", linetype = "longdash", size = 0.5, alpha = 0.8) +
  scale_x_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6), expand = c(0.05, 0)) +
  custom_theme + ylab("π") + xlab("Mbp") + facet_wrap(~chromosome, nrow = 5, scales="free")

# Second geom_rect specifically for chromosome 1. Filter data outside of aes().

width_inches <- 8
height_inches <- 10
dpi <- 800
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/04_figures/diversity_27_genomes_only_4_fold_with_ancestral_centr_for_poster_1Mb_centr.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(final_plot)
dev.off()

```
## Add rho
```{r}
```


# Create a boxplots
```{r}
data_full_range <- data_full_range %>%
  mutate(
    region = ifelse(window_start <= (start - 4000000) | window_start >= (stop + 4000000), "A", "P")
  )

my_colors <- c("P" = "gray30", "A" = "#527853")
# Group results per chromosome
results <- data_full_range %>% 
  dplyr::group_by(chromosome) %>%  do(tidy(wilcox.test(avg_pi_4_fold_window ~ region, data = .)))

# Adjust the p-values using BH correction
results$p.adj <- p.adjust(results$p.value, method = "fdr")
#Combine pvalue and start labels 

results <- results %>%
  mutate(signif_level = case_when(
    p.adj < 0.001 ~ '***',
    p.adj < 0.01  ~ '**',
    p.adj < 0.05  ~ '*',
    p.adj < 0.1   ~ '.',
    TRUE          ~ 'NS'  # NS indicates not significant
  ),
  label = paste("p.adj=", p.adj, " ", signif_level))



# Convert the adjusted p-values to scientific notation
options(digits=2)
results$p.adj.scientific <- format(results$p.adj, scientific = TRUE)


ymax <- max(data_full_range$avg_pi_4_fold_window)  # Get the max value to position labels 

boxplots <-ggboxplot(data_full_range, x = "region", y = "avg_pi_4_fold_window",
                  fill = "region", alpha = 0.6, color="region",notch=T,notchwidth = 0.8) + 
  geom_jitter(aes(col=region), size=2, show.legend = F,alpha = 0.6) +
   facet_wrap(~chromosome, ncol=5)+
  ylab("π") +
  xlab('region') + 
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) +
  scale_y_continuous(limits = c(0, 0.03)) +
  rotate_x_text(12) +
  theme(strip.text.x = element_text(size = 70)) +  # Adjust this value for desired size
  geom_text(data = results, aes(x = 1.5, y = 0.03, label = paste(signif_level)), size=8,col = "gray30") +  # Adjust the size value here for desired label size
  theme_minimal(base_size = 15) + 
  custom_theme +theme(legend.position="none")


print(boxplots)
width_inches <- 8
height_inches <- 6 
dpi <- 600
CairoPNG("eboxplots_diversity_27_thaliana_wo_ancestral_chr.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(boxplots)
dev.off()
```
# Boxplots including ancestral centromere 
```{r}
# Include ancestral stop for chromosome
data_full_range <- data_full_range %>%
  mutate(
    region = case_when(
      chromosome == "Chr1" & (window_start <= (start - 4000000) | window_start >= (ancestral_stop)) ~ "A",
      chromosome == "Chr1" & !(window_start <= (start - 4000000) | window_start >= (ancestral_stop)) ~ "P",
      chromosome != "Chr1" & (window_start <= (start - 4000000) | window_start >= (stop + 4000000)) ~ "A",
      chromosome != "Chr1" & !(window_start <= (start - 4000000) | window_start >= (stop + 4000000)) ~ "P"
    )
  )

my_colors <- c("P" = "gray30", "A" = "#527853")
# Group results per chromosome
results <- data_full_range %>% 
  dplyr::group_by(chromosome) %>%  do(tidy(wilcox.test(avg_pi_4_fold_window ~ region, data = .)))

# Adjust the p-values using BH correction
results$p.adj <- p.adjust(results$p.value, method = "fdr")
#Combine pvalue and start labels 

results <- results %>%
  mutate(signif_level = case_when(
    p.adj < 0.001 ~ '***',
    p.adj < 0.01  ~ '**',
    p.adj < 0.05  ~ '*',
    p.adj < 0.1   ~ 'NS',
    TRUE          ~ 'NS'  # NS indicates not significant
  ),
  label = paste("p.adj=", p.adj, " ", signif_level))



# Convert the adjusted p-values to scientific notation
options(digits=2)
results$p.adj.scientific <- format(results$p.adj, scientific = TRUE)


ymax <- max(data_full_range$avg_pi_4_fold_window)  # Get the max value to position labels 

boxplots <-ggboxplot(data_full_range, x = "region", y = "avg_pi_4_fold_window",
                  fill = "region", alpha = 0.6, color="region",notch=T,notchwidth = 0.8) + 
  geom_jitter(aes(col=region), size=2, show.legend = F,alpha = 0.6) +
   facet_wrap(~chromosome, ncol=5)+
  ylab("π") +
  xlab('region') + 
  scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) +
  scale_y_continuous(limits = c(0, 0.03)) +
  rotate_x_text(12) +
  theme(strip.text.x = element_text(size = 70)) +  # Adjust this value for desired size
  geom_text(data = results, aes(x = 1.5, y = 0.03, label = paste(signif_level)), size=6,col = "gray30") +  # Adjust the size value here for desired label size
  theme_minimal(base_size = 15) + 
  custom_theme +theme(legend.position="none")


print(boxplots)
width_inches <- 8
height_inches <- 6 
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/01_figures/boxplots_diversity_27_thaliana_with_ancestral_chr.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(boxplots)
dev.off()
```

## Additional code
# Generate plot Pi in 27 thaliana genomes with gene density
```{r}


ggplot(data_full_range, aes(x = window_start, y = gene_count)) +
    geom_col(fill="#776B5D",alpha=0.5) +
    scale_y_continuous(breaks=c(1, max_gene_count), limits=c(0, max_gene_count * 1.2)) +
    xlab(NULL) + ylab(NULL) +
    theme_minimal(base_size = 15) + custom_theme

#Extract theme minimal text
minimal_theme_settings <- ggplot2::theme_minimal(base_size = 15)$text
custom_theme <- theme(
  plot.background = element_blank(), 
  panel.background = element_blank(),
  panel.border = element_blank(), # Ensure no frame
  panel.grid = element_blank(),
  axis.title.x = element_blank(), 
  axis.ticks.x = element_blank(),
  axis.line = element_line(color = "gray30"), # Add this line to show x and y axis lines
  plot.margin = margin(0, 5, 0, 5),
  strip.background = element_blank(),
  strip.text.x = minimal_theme_settings
)


max_chrom_length <- max(data_full_range$window_start, na.rm = TRUE)
# Change NA to 0 
max_gene_count <- max(data_full_range$gene_count)
data_full_range <- lapply(data_full_range, function(x) replace(x, is.na(x), 0))
data_full_range <- as.data.frame(data_full_range)
#Calculate sd
data_full_range <- data_full_range %>%
  group_by(chromosome) %>%
  mutate(sd = sd(avg_pi_4_fold_window, na.rm = TRUE)) %>%
  ungroup() 

# Generate plot 
generate_chromosome_plots <- function(chr, show_legend=FALSE) {
  data_subset <- subset(data_full_range, chromosome == chr)
  data_subset$gene_count <- ifelse(is.na(data_subset$gene_count), 0, data_subset$gene_count)
  # Calculate means
  mean_avg_pi_4deg <- mean(data_full_range$avg_pi_4_fold_window, na.rm = TRUE)
  mean_avg_pi_0deg <- mean(data_full_range$avg_pi_0_fold_window, na.rm = TRUE)
  pi_plot <- ggplot(data_subset, aes(x = window_start, y = avg_pi_4_fold_window)) +
    geom_ribbon(aes(ymin = avg_pi_4_fold_window-sd, ymax = ifelse(avg_pi_4_fold_window > mean_avg_pi_4deg, avg_pi_4_fold_window+sd, avg_pi_4_fold_window-sd)), fill = "gray30", alpha = 0.5) +
    geom_vline(aes(xintercept = start-4000000), color = "gray30", linetype = "dashed", size=0.8) +
    #geom_vline(aes(xintercept = ancestral_start-4000000), color = "gray30", linetype = "dashed", size=0.8) +
     geom_vline(aes(xintercept = ancestral_stop), color = "gray30", linetype = "dashed", size=0.8) +
    geom_vline(aes(xintercept = stop+4000000), color = "gray30", linetype = "dashed", size=0.8) +
    geom_line(aes(y=avg_pi_4_fold_window), size=0.8,alpha=0.7, colour="#4B527E") +
    geom_hline(aes(yintercept = mean_avg_pi_4deg), color = "#4B527E", linetype = "dashed") +

    scale_x_continuous(labels = unit_format(unit = "M", scale = 1e-6), expand = c(0.05,0), limits = c(0, max_chrom_length)) +
    theme_minimal(base_size = 12) + custom_theme + xlab(NULL)+ylab(NULL) +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), panel.border = element_blank(),
          plot.margin = margin(0, 5, 0, 8))
  bar_plot <- ggplot(data_subset, aes(x = window_start, y = gene_count)) +
    geom_col(fill="#776B5D",alpha=0.5) +
    scale_x_continuous(labels = unit_format(unit = "M", scale = 1e-6), expand = c(0.05,0), limits = c(0, max_chrom_length))+
    scale_y_continuous(breaks=c(1, max_gene_count), limits=c(0, max_gene_count * 1.2)) +
    xlab(NULL) + ylab(NULL) +
    theme_minimal(base_size = 15) + custom_theme +
    theme(axis.title.x=element_blank(), panel.border = element_blank(), plot.margin = margin(1, 0, 0, 0))

  if (!show_legend) {
    bar_plot <- bar_plot + theme(legend.position = "none")
  }

  combined <- pi_plot / bar_plot
  return(combined + plot_layout(heights = c(7.5, 3.5)))
}

chromosomes <- unique(data_full_range$chromosome)

combined_plots <- lapply(chromosomes, function(chr) {
  show_legend <- chr == tail(chromosomes, n=1)
  generate_chromosome_plots(chr)
})

for(i in 1:length(combined_plots)) {
  combined_plots[[i]] <- combined_plots[[i]] + 
    theme(axis.title.y = element_blank(),
          axis.title.x = element_blank())
}

final_plot <- wrap_plots(combined_plots, ncol = 1) + theme_minimal() + custom_theme

grid.text("π", x = unit(0.01, "npc"), y = unit(0.5, "npc"), just = "center", rot = 90, gp = gpar(fontsize = 15, col = "gray30"))
grid.text("Mbp", x = unit(0.5, "npc"), y = unit(0.01, "npc"), just = "center", gp = gpar(fontsize = 15, col = "gray30"))

# Save the figure 

width_inches <- 8
height_inches <- 10
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/01_figures/diversity_27_genomes_only_4_fold_with_ancestral_centr.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)

print(final_plot)

grid.text("π", x = unit(0.01, "npc"), y = unit(0.5, "npc"), just = "center", rot = 90, gp = gpar(fontsize = 15, col ="gray30"))
grid.text("Mbp", x = unit(0.5, "npc"), y = unit(0.01, "npc"), just = "center", gp = gpar(fontsize = 15, col = "gray30"))

dev.off()

```

```{r}
gene_count <- ggplot(data_full_range, aes(window_start, gene_count))+geom_bar(stat="identity", fill="#A0937D")+facet_wrap(~chromosome, ncol=1, scales="free") +custom_theme
width_inches <- 15
height_inches <- 10
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/01_figures/gene_count_27_genomes.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(gene_count)
dev.off()

```




