---
title: "analysis_of_syntenic_genes"
format: pdf
editor: visual
autor: Lisa Grigoreva 18/02/26 
---

## Genetic neighborhood influences pi changes

We test the hypothesis that single-gene nucleotide diversity adapts to the local neighborhood. Implying that single-gene diversity level reflects the neighborhood in which the gene currently lives, not the ancestral neighborhood. 

We align 31 *A.lyrata* genomes and 27 *A.thaliana* genomes to their corresponding reference genomes (MN47 and TAIR10) using pannagram. We estimate diversity in 4fold (synonymous) sites.

Ortholgous genes calculated by running simsearch in pannagram between TAIR10 and MN47. Genes with \>85% identity were considered as syntenic.

## Calculate how many genes moved together

```{r}
#| echo: false
library(dplyr)
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
# Read pannagram genes
pannagram_1_1_genes <- read.table("/groups/nordborg/projects/genome_organization/01_data/tair10/1_1_orth_genes_pannagram.txt", header=T)
# Select necessary columns 
pannagram_1_1_genes <- pannagram_1_1_genes %>% dplyr::select(chrom_thal,gene_thal,start_thal,end_thal,start_lyr,end_lyr,chrom_lyr)
```

## Dotplot for transposed genes for all chromosomes

```{r}
# Chromosome lengths and cumulative positions for A. thaliana
pannagram_1_1_genes$chrom_lyr <- gsub("scaffold_", "", pannagram_1_1_genes$chrom_lyr)

thal_lengths <- pannagram_1_1_genes %>%
  group_by(chrom_thal) %>%
  dplyr::summarise(max_pos = max(end_thal)) %>%
  arrange(chrom_thal) %>%
  mutate(cum_start = lag(cumsum(max_pos), default = 0))

# Chromosome lengths and cumulative positions for A. lyrata
lyr_lengths <- pannagram_1_1_genes %>%
  group_by(chrom_lyr) %>%
  dplyr::summarise(max_pos = max(end_lyr)) %>%
  arrange(chrom_lyr) %>%
  mutate(cum_start = lag(cumsum(max_pos), default = 0))

# Add cumulative positions to the main df
pannagram_1_1_genes <- pannagram_1_1_genes %>%
  left_join(thal_lengths %>% select(chrom_thal, cum_start), by = "chrom_thal") %>%
  dplyr::rename(cum_start_thal = cum_start) %>%
  left_join(lyr_lengths %>% select(chrom_lyr, cum_start), by = "chrom_lyr") %>%
  dplyr::rename(cum_start_lyr = cum_start) %>%
  dplyr::mutate(
    cum_pos_thal = cum_start_thal + (start_thal + end_thal) / 2,
    cum_pos_lyr = cum_start_lyr + (start_lyr + end_lyr) / 2
  )
# plot
pannagram_1_1_genes$chrom_lyr <- gsub("_", "", pannagram_1_1_genes$chrom_lyr)
pannagram_1_1_genes$chrom_lyr <- gsub("scaffold", "", pannagram_1_1_genes$chrom_lyr)
ggplot(pannagram_1_1_genes, aes(x = cum_pos_lyr, y = cum_pos_thal, color = as.factor(chrom_lyr))) +
  geom_point(alpha = 0.6, size = 2) +
  geom_hline(data = thal_lengths, aes(yintercept = cumsum(max_pos)), 
             linetype = "dashed", color = "gray30", linewidth = 0.1) +
  scale_x_continuous(labels = function(x) x / 1e6) +
  scale_y_continuous(labels = function(x) x / 1e6) +
  scale_color_manual(values = c("1" = "#FFFF38", 
                                 "2" = "#E41A1C",
                                 "3" = "blue",
                                 "4" = "#F781BF",
                                 "5" = "orange",
                                 "6" = "darkgreen",
                                 "7" = "lightblue",
                                 "8" = "lightpink")) +
  labs(
    x = "A. lyrata position (Mbp)",
    y = "A. thaliana position (Mbp)",
    color = "A. lyrata\nchromosome"
  ) +
  theme_bw(base_size  = 16) +
  theme(legend.position = "right")

```

We search for genes which appear to be transposed (single dots on the dotpot).

The search works in the following way:

For each syntenic gene we create a window around its position in A. thaliana, then we search genes using the following logic:

Are there other genes from the **same *A. lyrata* chromosome** within my ***A. thaliana*** **window** that are at least window size away?

The gene is classified as transposed if:

1.  It is isolated. No other genes from the same *A. lyrata* chromosome are found nearby in the *A. thaliana* genome (within 300 kb, no genes)
2.  It can have syntenic neighbors in *A. lyrata* meaning that the gene comes from the syntenic region (single moves without synthetic neighbors are also considered)

```{r}
identify_transposed_genes <- function(data, window_size = 250000, fixed_window_size = 500000) {
  data <- data %>%
    arrange(chrom_thal, start_thal) %>%
    mutate(
      # Gene-centered windows (existing)
      window_start_thal = pmax(0, start_thal - window_size),
      window_end_thal   = start_thal + window_size,        
      window_start_lyr  = pmax(0, start_lyr - window_size),
      window_end_lyr    = start_lyr + window_size,        
      cumulative_window_start_thal = pmax(0, cum_pos_thal - window_size),
      cumulative_window_end_thal   = cum_pos_thal + window_size,
      cumulative_window_start_lyr  = pmax(0, cum_pos_lyr - window_size),
      cumulative_window_end_lyr    = cum_pos_lyr + window_size,
      # Fixed windows (new) 
      # Assigns each gene to a fixed bin: 0-500000, 500000-1000000, etc.
      fixed_window_start_thal = floor(start_thal / fixed_window_size) * fixed_window_size,
      fixed_window_end_thal   = fixed_window_start_thal + fixed_window_size,
      fixed_window_start_lyr  = floor(start_lyr  / fixed_window_size) * fixed_window_size,
      fixed_window_end_lyr    = fixed_window_start_lyr  + fixed_window_size
    ) %>%
    mutate(
      has_syntenic_neighbors_thal = sapply(1:n(), function(i) {
        sum(
          chrom_thal == chrom_thal[i] &
          chrom_lyr  == chrom_lyr[i] &
          start_thal >= window_start_thal[i] &
          start_thal <= window_end_thal[i] &
          seq_len(n()) != i
        ) > 0
      }),
      
      has_syntenic_neighbors_lyr = sapply(1:n(), function(i) {
        sum(
          chrom_lyr  == chrom_lyr[i] &
          chrom_thal == chrom_thal[i] &
          start_lyr >= window_start_lyr[i] &
          start_lyr <= window_end_lyr[i] &
          seq_len(n()) != i
        ) > 0
      }),
      
      is_transposed = !has_syntenic_neighbors_thal
    )
  
  return(data)
}

pannagram_1_1_genes <- identify_transposed_genes(pannagram_1_1_genes, 
                                                  window_size = 250000, 
                                                  fixed_window_size = 500000)
```

```{r}
# Fixed window gene counts - THAL
genes_per_fixed_window_thal <- pannagram_1_1_genes %>%
  group_by(chrom_thal, fixed_window_start_thal, fixed_window_end_thal) %>%
  summarise(n_genes_thal = n_distinct(gene_thal), .groups = "drop") %>%
  arrange(chrom_thal, fixed_window_start_thal)

# Fixed window gene counts - LYR
genes_per_fixed_window_lyr <- pannagram_1_1_genes %>%
  group_by(chrom_lyr, fixed_window_start_lyr, fixed_window_end_lyr) %>%
  summarise(n_genes_lyr = n_distinct(gene_thal), .groups = "drop") %>%
  arrange(chrom_lyr, fixed_window_start_lyr)
```

Transposed genes with their relative position

```{r}
ggplot(pannagram_1_1_genes, aes(x = cum_pos_lyr, y = cum_pos_thal)) +
  geom_hline(data = thal_lengths, aes(yintercept = cumsum(max_pos)), 
             linetype = "dashed", color = "gray30", linewidth = 0.3) +
  
  # Plot non-transposed genes in gray
  geom_point(data = subset(pannagram_1_1_genes, !is_transposed),
             shape = 21, fill = NA, color = "gray70",
             alpha = 0.3, size = 2, stroke = 0.8) +
  
  # Plot transposed genes in colorblind-friendly color
  geom_point(data = subset(pannagram_1_1_genes, is_transposed),
             shape = 21, fill = "#5C5470", color = "black", 
             size = 2.5, alpha = 0.9, stroke = 1.5) +
  
  scale_x_continuous(labels = function(x) x / 1e6) +
  scale_y_continuous(labels = function(x) x / 1e6) +
  
  labs(
    x = "A. lyrata position (Mbp)",
    y = "A. thaliana position (Mbp)"
  ) +
  theme_bw(base_size = 16) +
  theme(legend.position = "none")

```

We calculate nucleotide diversity in the corresponding windows using synonymous sites.

```{r}
# Sploit genes in high-high , high-low and see correlation 
# How z scores are correlated? How jumped dots are correlated?  
# Without z score, how they are correlated? 

library(data.table)
# Read pi in synonymous sites 
pi_4fold_thal <- fread("/groups/nordborg/projects/genome_organization/manuscript/04_output/002_div_thal_lyr/4fol_sites_thal_27g_tair10_based_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
pi_4fold_lyr <- fread("/groups/nordborg/projects/genome_organization/manuscript/04_output/002_div_thal_lyr/4fol_sites_lyr_31g_mn47_based_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
genes_athaliana <- fread("/groups/nordborg/projects/ddm1_resilencing/11_resources/ref_genome/TAIR10_GFF3_genes_transposons.gff")  %>% filter(V3=="gene")
```

```{r}
library(data.table)
# Convert to data.table for speed
genes_dt <- as.data.table(pannagram_1_1_genes)
pi_thal_dt <- as.data.table(pi_4fold_thal)
pi_lyr_dt <- as.data.table(pi_4fold_lyr)

# Add gene IDs
genes_dt[, gene_id := .I]

# THALIANA - Gene pi
genes_dt[, chr_thal := paste0("Chr", chrom_thal)]
setkey(pi_thal_dt, chromosome, window_pos_1)

gene_pi_thal <- pi_thal_dt[genes_dt, 
  on = .(chromosome = chr_thal, 
         window_pos_1 >= start_thal, 
         window_pos_1 <= end_thal),
  .(count_diffs = sum(count_diffs), 
    count_comparisons = sum(count_comparisons)),
  by = .EACHI]

gene_pi_thal[, pi_gene_thal := count_diffs / count_comparisons]

# THALIANA - Window pi
window_pi_thal <- pi_thal_dt[genes_dt,
  on = .(chromosome = chr_thal,
         window_pos_1 >= window_start_thal,
         window_pos_1 <= window_end_thal),
  .(count_diffs = sum(count_diffs),
    count_comparisons = sum(count_comparisons),
    n_sites_window_thal = .N),
  by = .EACHI]

window_pi_thal[, pi_window_thal := count_diffs / count_comparisons]

# LYRATA - Gene pi
genes_dt[, chr_lyr := paste0("Chr", chrom_lyr)]

gene_pi_lyr <- pi_lyr_dt[genes_dt,
  on = .(chromosome = chr_lyr,
         window_pos_1 >= start_lyr,
         window_pos_1 <= end_lyr),
  .(count_diffs = sum(count_diffs),
    count_comparisons = sum(count_comparisons)),
  by = .EACHI]

gene_pi_lyr[, pi_gene_lyr := count_diffs / count_comparisons]

# LYRATA - Window pi
window_pi_lyr <- pi_lyr_dt[genes_dt,
  on = .(chromosome = chr_lyr,
         window_pos_1 >=window_start_lyr,
         window_pos_1 <= window_end_lyr),
  .(count_diffs = sum(count_diffs),
    count_comparisons = sum(count_comparisons),
   n_sites_window_lyr = .N),
  by = .EACHI]

window_pi_lyr[, pi_window_lyr := count_diffs / count_comparisons]

# Add results back
pannagram_1_1_genes$pi_gene_thal <- gene_pi_thal$pi_gene_thal
pannagram_1_1_genes$pi_window_thal <- window_pi_thal$pi_window_thal
pannagram_1_1_genes$n_sites_window_thal <- window_pi_thal$n_sites_window_thal
pannagram_1_1_genes$pi_gene_lyr <- gene_pi_lyr$pi_gene_lyr
pannagram_1_1_genes$pi_window_lyr <- window_pi_lyr$pi_window_lyr
pannagram_1_1_genes$n_sites_window_lyr<- window_pi_lyr$n_sites_window_lyr
```

Where are transposed genes in the genome of A. thaliana?

```{r}
# chromosome boundary positions
library(ggplot2)
library(patchwork)
library(dplyr)

library(ggplot2)
library(patchwork)
library(dplyr)

df <- pannagram_1_1_genes %>% filter(is_transposed == TRUE)

lyr_bounds  <- df %>% group_by(chrom_lyr)  %>% summarise(x = max(cum_pos_lyr)  / 1e6)
thal_bounds <- df %>% group_by(chrom_thal) %>% summarise(x = max(cum_pos_thal) / 1e6)

med_lyr  <- median(df$pi_window_lyr,  na.rm = TRUE)
med_thal <- median(df$pi_window_thal, na.rm = TRUE)

# ── TOP: lyrata ───────────────────────────────────────────────────────────────
p_lyr <- df %>%
  arrange(cum_pos_lyr) %>%
  ggplot(aes(x = cum_pos_lyr / 1e6, y = pi_window_lyr)) +
  geom_vline(data = lyr_bounds, aes(xintercept = x), linetype = "dashed", color = "gray50", linewidth = 0.3) +
  geom_ribbon(aes(ymin = med_lyr, ymax = pmax(pi_window_lyr, med_lyr)),
              fill = "#285A48", alpha = 0.3) +
  geom_line(color = "#285A48", linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = med_lyr, color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = Inf, y = med_lyr, label = paste0("", round(med_lyr, 4)),
           hjust = 1.1, vjust = -0.5, color = "black", size = 3.5) +
  xlab(NULL) + ylab("π in A. lyrata") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

# ── BOTTOM: thaliana ──────────────────────────────────────────────────────────
p_thal <- df %>%
  arrange(cum_pos_thal) %>%
  ggplot(aes(x = cum_pos_thal / 1e6, y = pi_window_thal)) +
  geom_vline(data = thal_bounds, aes(xintercept = x), linetype = "dashed", color = "gray50", linewidth = 0.3) +
  geom_ribbon(aes(ymin = med_thal, ymax = pmax(pi_window_thal, med_thal)),
              fill = "#285A48", alpha = 0.3) +
  geom_line(color = "#285A48", linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = med_thal, color = "black", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = Inf, y = med_thal, label = paste0(" ", round(med_thal, 4)),
           hjust = 1.1, vjust = -0.5, color = "black", size = 3.5) +
  xlab("Cumulative position (Mbp)") + ylab("π in A. thaliana") +
  theme_bw()

p_lyr / p_thal
```

Distance between transposed genes in A. thaliana genome

```{r}
# Calculate distance to next gene on the same chromosome
pannagram_1_1_genes <- pannagram_1_1_genes %>%
  arrange(chrom_thal, start_thal) %>%
  group_by(chrom_thal) %>%
  mutate(
    next_gene_start = lead(start_thal),
    distance_to_next_gene_thal = next_gene_start - end_thal
  ) %>%
  ungroup()

pannagram_1_1_genes %>% 
  filter(is_transposed == "TRUE") %>% 
  ggplot(aes(x = window_start_thal, y = pi_window_thal)) +
  geom_point(aes(color = distance_to_next_gene_thal)) +
  theme_minimal()+facet_wrap(~chrom_thal, ncol=1)
```

Windows

```{r}
pannagram_1_1_genes$distanse_win <- pannagram_1_1_genes$window_end_thal-pannagram_1_1_genes$window_start_thal
hist(pannagram_1_1_genes$distanse_win)
```

```{r}
# Genes per window for THAL
# Genes per window for THAL
hist(pannagram_1_1_genes$n_sites_window_thal)
hist(pannagram_1_1_genes$n_sites_window_lyr)

pannagram_1_1_genes <- pannagram_1_1_genes[pannagram_1_1_genes$n_sites_window_thal>5000 & pannagram_1_1_genes$n_sites_window_lyr > 5000,]

# Calculate how many genes per window 
genes_athaliana$gene_thal <- gsub(".*;Name=", "", genes_athaliana$V9)
genes_athaliana$chrom_thal <- as.numeric(gsub("Chr", "",genes_athaliana$V1 ))

genes_in_windows <- pannagram_1_1_genes %>%
  left_join(genes_athaliana, by = join_by(chrom_thal)) %>%
  filter(V4 >= window_start_thal & V5 <= window_end_thal,)

# Count genes per window
gene_counts <- genes_in_windows %>%
  group_by(window_start_thal, chrom_thal) %>%
  summarise(n_genes = n(), .groups = "drop")

pannagram_1_1_genes <- left_join(pannagram_1_1_genes, gene_counts, 
                                  by = c("window_start_thal", "chrom_thal"))

pannagram_1_1_genes_transposed_counts <-pannagram_1_1_genes[pannagram_1_1_genes$is_transposed=="TRUE",]
pannagram_1_1_genes_transposed_counts$n_genes
```

Then we can add diversity levels to the x and y axes to see visually if the single moved genes came from the same change or not their diversity neighborhood

```{r}

safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

median_pi_thal <- median(pannagram_1_1_genes$pi_window_thal, na.rm = TRUE)
median_pi_lyr <- median(pannagram_1_1_genes$pi_window_lyr, na.rm = TRUE)

# Get axis limits from data
y_limits <- range(pannagram_1_1_genes$cum_pos_thal)
x_limits <- range(pannagram_1_1_genes$cum_pos_lyr)

# Identify regions above median
high_pi_thal <- pannagram_1_1_genes %>%
  filter(pi_window_thal > median_pi_thal)

high_pi_lyr <- pannagram_1_1_genes %>%
  filter(pi_window_lyr > median_pi_lyr)

# Main dot plot
p_main <- ggplot(pannagram_1_1_genes, aes(x = cum_pos_lyr, y = cum_pos_thal)) +
  geom_hline(data = thal_lengths, aes(yintercept = cumsum(max_pos)), 
             linetype = "solid", color = "gray30", linewidth = 1) +
  # Add vertical lines for high π lyrata
  geom_vline(data = high_pi_lyr, aes(xintercept = cum_pos_lyr),
             color = "#CC6677", alpha = 0.02, linewidth = 0.5) +
  # Add horizontal lines for high π thaliana
  geom_hline(data = high_pi_thal, aes(yintercept = cum_pos_thal),
             color = "#117733", alpha = 0.02, linewidth = 0.5) +
  geom_point(data = subset(pannagram_1_1_genes, !is_transposed),
             shape = 21, fill = NA, color = "#999999",
             alpha = 0.3, size = 2, stroke = 0.8) +
  geom_point(data = subset(pannagram_1_1_genes, is_transposed),
             shape = 21, fill = "gray30", color = "gray30", 
             size = 2.5, alpha = 0.9, stroke = 1.5) +
  scale_x_continuous(labels = function(x) x / 1e6, limits = x_limits) +
  scale_y_continuous(labels = function(x) x / 1e6, limits = y_limits) +
  labs(x = "A. lyrata position (Mbp)", y = "A. thaliana position (Mbp)") +
  theme_bw(base_size = 16) +
  theme(legend.position = "none",
        plot.margin = margin(5, 5, 5, 5))

# Right plot - Thaliana diversity (aligned with y-axis)
p_right <- ggplot(pannagram_1_1_genes, aes(x = cumulative_window_start_thal, y = pi_window_thal)) +
  geom_ribbon(aes(ymin = median_pi_thal, 
                  ymax = pmax(pi_window_thal, median_pi_thal)),
              fill = "#117733", alpha = 0.5) +
  geom_line(color = "#117733", linewidth = 1, alpha = 0.5) +
  geom_hline(yintercept = median_pi_thal, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_x_continuous(limits = y_limits) +  # Match main plot y-axis
  labs(x = "π Thal", y = "") +
  theme_bw(base_size = 12) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.margin = margin(5, 0, 5, 5)) +
  coord_flip()

# Top plot - Lyrata diversity (aligned with x-axis)
p_top <- ggplot(pannagram_1_1_genes, aes(x = cumulative_window_start_lyr, y = pi_window_lyr)) +
  geom_ribbon(aes(ymin = median_pi_lyr, 
                  ymax = pmax(pi_window_lyr, median_pi_lyr)),
              fill = "#CC6677", alpha = 0.5) +
  geom_line(color = "#CC6677", linewidth = 1, alpha = 0.5) +
  geom_hline(yintercept = median_pi_lyr, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_x_continuous(limits = x_limits) +  # Match main plot x-axis
  labs(y = "π Lyr", x = "") +
  theme_bw(base_size = 12) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = margin(5, 5, 0, 5))

# Combine with patchwork
library(patchwork)
p_top + plot_spacer() + p_main + p_right + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))
```

We distribute these genes in the following categories:

-   High diversity of the neighborhood in *A. thaliana* and high diversity of the gene ( relative to the mean in A. thaliana) -\> gene lives in the high diversity neighborhood

-   Low diversity of the neighborhood in *A. thaliana* and low diversity of the gene ( relative to the mean in A. thaliana) -\> gene lives in the low diversity neighborhood

    We do the same with *A. lyrata*

-   High diversity of the neighborhood in *A. lyrata* and high diversity of the gene ( relative to the mean in *A. lyrata*) -\> gene lives in the high diversity neighborhood

-   Low diversity of the neighborhood in *A. lyrata* and low diversity of the gene ( relative to the mean in *A. lyrata*) -\> gene lives in the low diversity neighborhood

```{r}
library(tidyr)
median_pi_thal <- median(pannagram_1_1_genes$pi_window_thal, na.rm = TRUE)
median_pi_lyr  <- median(pannagram_1_1_genes$pi_window_lyr,  na.rm = TRUE)

transposed_genes_combined <- pannagram_1_1_genes %>%
  mutate(
    context_thal = case_when(
      pi_window_thal > median_pi_thal ~ "high_thal",
      pi_window_thal < median_pi_thal ~ "low_thal",
      pi_window_thal == median_pi_thal ~ "median_thal"
    ),
    context_lyr = case_when(
      pi_window_lyr > median_pi_lyr ~ "high_lyr",
      pi_window_lyr < median_pi_lyr ~ "low_lyr",
      pi_window_lyr == median_pi_lyr ~ "median_lyr"
    ),
    context_combined = case_when(
      pi_window_thal > median_pi_thal & pi_window_lyr > median_pi_lyr ~ "high_high",
      pi_window_thal > median_pi_thal & pi_window_lyr < median_pi_lyr ~ "high_low",
      pi_window_thal < median_pi_thal & pi_window_lyr > median_pi_lyr ~ "low_high",
      pi_window_thal < median_pi_thal & pi_window_lyr < median_pi_lyr ~ "low_low",
      TRUE ~ "median"  # catches any exact median ties
    )
  )

plot_data <- transposed_genes_combined %>%
  mutate(
    # Z-scale within each species separately
    # thal: scale gene pi relative to all thal genes
    scaled_pi_thal_gene   = (pi_gene_thal   - mean(pi_gene_thal,   na.rm=TRUE)) / sd(pi_gene_thal,   na.rm=TRUE),
    scaled_pi_thal_neighb = (pi_window_thal - mean(pi_window_thal, na.rm=TRUE)) / sd(pi_window_thal, na.rm=TRUE),
    # lyr: scale gene pi relative to all lyr genes
    scaled_pi_lyr_gene    = (pi_gene_lyr    - mean(pi_gene_lyr,    na.rm=TRUE)) / sd(pi_gene_lyr,    na.rm=TRUE),
    scaled_pi_lyr_neighb  = (pi_window_lyr  - mean(pi_window_lyr,  na.rm=TRUE)) / sd(pi_window_lyr,  na.rm=TRUE),
    # Difference between neighbourhoods
    lyr_neighb_thal_neighb = scaled_pi_lyr_neighb - scaled_pi_thal_neighb,
    # Difference between genes
    lyr_gene_thal_gene     = scaled_pi_lyr_gene   - scaled_pi_thal_gene,
    # Gene in new environment vs new neighbourhood
    lyr_gene_thal_neighb   =  scaled_pi_lyr_gene - scaled_pi_thal_neighb
  )
# plot 
plot_data %>%
  filter(is_transposed == TRUE) %>%
  pivot_longer(
    cols = c(lyr_gene_thal_gene, lyr_neighb_thal_neighb,lyr_gene_thal_neighb),
    names_to = "environment",
    values_to = "delta_pi_scaled"
  ) %>%
  ggplot(aes(x = environment, y = delta_pi_scaled, fill = context_combined)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.8, outlier.shape = NA) +
  scale_x_discrete(labels = c(
    "delta_pi_scaled_old_env" = "Old Environment",
    "delta_pi_scaled_new_env" = "New Environment"
  )) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Environment",
    y = "Scaled Delta Pi",
    fill = "Context",
    title = ""
  ) +
  theme_bw()


```

```{r}
library(ggpubr)

plot_data %>%
  filter(is_transposed == FALSE) %>%
  ggplot(aes(pi_gene_lyr, pi_gene_thal)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top",
         aes(label = after_stat(r.label))) +xlab("gene pi A.lyrata") + ylab("gene pi A. thaliana")+theme_bw(base_size = 16)+facet_wrap(~context_combined, scales = "free")

```

Hypothesis of gene moves from the one context to the other:

1.  Genes that have moved will now match its current “diversity environment” simply as a consequence of background selection.

2.  Genes that move to a different “diversity environment” will be selected against, and hence under-represented relative to genes that moved to the same environment and that were not thusly selected.

    To test this hypothesis we show the enrichment of the transposed genes in the 4 categories:

    -   gene is from high environment in A.lyrata ended up in high environment in A. thaliana

    -   gene is from low environment in A.lyrata ended up in low environment in A. thaliana

        These two categories would be enriched. The following two expected to be underrepresented

    -   gene is from low environment in A.lyrata ended up in high environment in A. thaliana

    -   gene is from high environment in A.lyrata ended up in low environment in A. thaliana

To test this, we create a contingency table with transposed genes according to their categories and percentage of the aligned genes which are in high or low category

Then we do a chisquare test?

```{r}
# Calculate proportion of the genes in high thal and high lyr
n_total <- nrow(plot_data)

# p = proportion of genes in high neighbourhood
p_thal <- sum(plot_data$context_thal == "high_thal", na.rm = TRUE) / n_total
p_lyr  <- sum(plot_data$context_lyr  == "high_lyr",  na.rm = TRUE) / n_total

# Expected proportions under independence
expected_high_high <- p_thal * p_lyr
expected_low_low   <- (1 - p_thal) * (1 - p_lyr)
expected_high_low  <- p_thal * (1 - p_lyr)
expected_low_high  <- (1 - p_thal) * p_lyr

# Sanity check - should sum to 1
cat("Sum of expected:", expected_high_high + expected_low_low + 
                        expected_high_low  + expected_low_high, "\n")

# Observed in transposed genes
transposed_only <- plot_data %>%
  filter(is_transposed == TRUE)
n_transposed <- nrow(transposed_only)

comparison <- data.frame(
  transition    = c("high_high", "low_low", "high_low", "low_high"),
  observed_n    = c(
    sum(transposed_only$context_combined == "high_high"),
    sum(transposed_only$context_combined == "low_low"),
    sum(transposed_only$context_combined == "high_low"),
    sum(transposed_only$context_combined == "low_high")
  ),
  expected_prop = c(expected_high_high, expected_low_low, 
                    expected_high_low,  expected_low_high)
) %>%
  mutate(
    observed_prop = observed_n / n_transposed,
    expected_n    = expected_prop * n_transposed,
    enrichment    = observed_prop / expected_prop,
    environment   = ifelse(transition %in% c("high_high", "low_low"), 
                           "same", "different")
  )

print(comparison)

# Chi-square with 4 categories now
chisq_hw <- chisq.test(
  x = comparison$observed_n,
  p = comparison$expected_prop
)
print(chisq_hw)
```

In which categories transposed genes are enriched?

```{r}
# Pivot to long format for plotting
comparison_long <- comparison %>%
  pivot_longer(
    cols = c(observed_prop, expected_prop),
    names_to = "type",
    values_to = "proportion"
  ) %>%
  mutate(
    type = recode(type,
      "observed_prop" = "Observed",
      "expected_prop" = "Expected"
    )
  )

ggplot(comparison_long, aes(x = transition, y = proportion, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = paste0("n = ", round(proportion * n_transposed))),
            position = position_dodge(width = 0.6),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Observed" = "#E69F00", 
                                "Expected"  = "#56B4E9")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
  labs(
    x    = "Context",
    y    = "Proportion",
    fill = NULL,
  ) +
  theme_bw() 

```

Moved VS not moved genes. Is the transposition is the context associated with transposition?

Is the distribution across diversity contexts different between transposed and non-transposed genes?

Null hypothesis:Transposed genes follow genome-wide background rates

```{r}
# Build contingency table: transposed vs non-transposed across 4 contexts
contingency <- plot_data %>%
  filter(!is.na(context_combined)) %>%
  count(is_transposed, context_combined) %>%
  pivot_wider(names_from = context_combined, values_from = n, values_fill = 0)

print(contingency)

# Chi-square test of independence: is transposition distribution
# different between transposed and non-transposed genes?
chisq_contingency <- chisq.test(
  table(plot_data$is_transposed, plot_data$context_combined)
)
print(chisq_contingency)

# Enrichment: observed / expected per group
enrich_table <- as.data.frame(chisq_contingency$observed) %>%
  rename(observed = Freq) %>%
  mutate(
    expected   = as.vector(chisq_contingency$expected),
    enrichment = observed / expected
  )

chisq_contingency


```

```{r}
enrich_long <- enrich_table %>%
  rename(transition = Var2,
         is_transposed = Var1) %>%  
  group_by(is_transposed) %>%
  mutate(proportion = observed / sum(observed)) %>%
  ungroup()

ggplot(enrich_long, aes(x = transition, y = proportion, fill = is_transposed)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = paste0("n = ", observed)),
            position = position_dodge(width = 0.6),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9"),
                    labels = c("TRUE" = "Transposed", "FALSE" = "Non-transposed")) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.15))) +
  labs(
    x    = "Context",
    y    = "Proportion of Genes",
    fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

Enrichment for the transposed genes

```{r}
write.table(plot_data, "/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/syntenic_and_transposed_genes.txt", quote = F, row.names = F)
```
