---
title: "rho_visualization"
output: html_document
date: "2024-04-07"
---

```{r setup, include=FALSE}
library(data.table)
library(tidyr)

# Set the working directory to where your files are located
setwd("/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr5/") # Update this with your actual directory path

# List all files that match your criteria (assuming 'res.txt' is part of all filenames)
files <- list.files(pattern="res\\.txt$")

# Initialize an empty data table to store the combined data
combined_data <- rbindlist(lapply(files, function(file) {
  # Read the current file
  dt <- fread(file)
  
  # Remove the first line of data after the header
  if (nrow(dt) > 1) {
    dt <- dt[-1, ]
  }
  
  # Check if 'Loci' column exists and multiply its values by 1000
  if ("Loci" %in% names(dt)) {
    dt[, Loci := Loci * 1000]
  }
  
  return(dt)
}), use.names = TRUE, fill = TRUE)

# Optional: Save the combined data to a new file
fwrite(combined_data, "combined_ldhat_output_with_loci_adjusted.csv", sep = "\t")
print("Combined LDhat output with Loci adjusted and without the first data row of each file has been saved.")

```
# Adjust per window 
```{r}
# Calculate window limits
window_size <- 500000
overlap_size <- 5000
mid_overlap <- 2500 # Middle of the overlap

# Function to adjust Mean_rho within overlaps
adjust_overlaps <- function(dt) {
  # Iterate over the windows, skipping the first since it has no previous overlap
  for (start_loci in seq(from = window_size, to = max(dt$Loci), by = window_size)) {
    # Define the overlap range
    overlap_start <- start_loci - overlap_size + mid_overlap + 1
    overlap_end <- start_loci + mid_overlap
    
    # Subset data for the overlap
    overlap_dt <- dt[Loci >= overlap_start & Loci <= overlap_end]
    
    # Check if there is overlap data to process
    if (nrow(overlap_dt) > 0) {
      # Calculate the mean of Mean_rho for the overlap
      mean_rho <- mean(overlap_dt$Mean_rho)
      
      # Assign the averaged Mean_rho back to the main dataset for the overlap range
      dt[Loci >= overlap_start & Loci <= overlap_end, Mean_rho := mean_rho]
    }
  }
  
  return(dt)
}

# Apply the adjustment function to your dataset
combined_data_adjusted <- adjust_overlaps(combined_data)

# Now combined_data_adjusted contains Mean_rho averaged over the middle 2,500 bp of each overlap

```
# Visualize
```{r}
# Define the window size
# Define window size and step size
# Define window size and parameters
window_size <- 500000
step_size <- window_size  # Non-overlapping windows for efficiency
min_loci <- 1
max_loci <- 26975502

# Pre-calculate number of windows needed
n_windows <- ceiling((max_loci - min_loci + 1) / window_size)

# Pre-allocate result data.table for better performance
mean_rho_per_window <- data.table(
  Start = numeric(n_windows),
  End = numeric(n_windows),
  Mean_rho_mean = numeric(n_windows)
)

# Sort data by Loci first (if not already sorted)
setorder(combined_data_adjusted, Loci)

# Create a more efficient windowing function
calculate_windows <- function(dt) {
  # Add window index to each row
  dt[, window_idx := floor((Loci - min_loci) / window_size) + 1]
  
  # Calculate means by window index
  window_means <- dt[, .(
    Start = min_loci + (window_idx - 1) * window_size,
    End = min_loci + window_idx * window_size - 1,
    Mean_rho_mean = mean(Mean_rho)
  ), by = window_idx]
  
  # Remove window_idx column
  window_means[, window_idx := NULL]
  return(window_means)
}

# Calculate windows efficiently
mean_rho_per_window <- calculate_windows(combined_data_adjusted)

# Add chromosome information
mean_rho_per_window[, chromosome := "Chr5"]


# Save the results to a file
fwrite(mean_rho_per_window, "/groups/nordborg/projects/genome_organization/08_manuscripts/03_results/historical_recombination/chr5_ldhat_output_with_loci_adjusted_windows_500kb.csv", sep = "\t")
```

```

```{r}
ggplot(mean_rho_per_window, aes(x = Start, y = Mean_rho_mean)) +
  geom_point(size = 0.8, alpha = 0.7, colour = "#527853") +
  geom_line(size = 0.8, alpha = 0.7, colour = "#527853") +
  custom_theme + ylab("rho") + xlab("Mbp")

mean_rho_per_window$chromosome <- "Chr1"
```
# Upload pi 
```{r}
pi_4fold_thal <- read.table("/groups/nordborg/projects/genome_organization/03_processing/001_pi_pacbio_27g_thaliana/pixy_pi_thaliana_genic_and_non_genic/all_chr_biall_genic_4_fold_all_genes_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()

pi_0fold_thal <- read.table("/groups/nordborg/projects/genome_organization/03_processing/001_pi_pacbio_27g_thaliana/pixy_pi_thaliana_genic_and_non_genic/all_chr_biall_genic_4_fold_all_genes_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()


#pi_0fold_thal <- read.table("/groups/nordborg/projects/genome_organization/03_processing/001_pi_pacbio_27g_thaliana/pixy_pi_thaliana_genic_and_non_genic/all_chr_biall_genic_0_fold_all_genes_pi.txt", header=TRUE, sep='\t') %>% na.omit() %>% distinct()
centromeres <- read.csv2('/groups/nordborg/projects/genome_organization/01_data/tair10/centr_bound.txt',header=T, sep=",")
chromosome_lengths <- data.frame(
  chromosome = c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"),
  Golden_path_length = c("30427671", "19698289", "23459830", "18585056", "26975502")
)
# Full join of 0 and 4 fold degenerate sites
pi_combined <- full_join(pi_4fold_thal, pi_0fold_thal, 
                         by=c("chromosome", "window_pos_1", "window_pos_2"), 
                         suffix=c("_4fold", "_0fold")) %>%
  mutate(window_start = (window_pos_1 %/% 500000) * 500000 + 1,
         window_end = window_start + 499999)

# Calculate gene density
gene_density <- pi_combined %>%
  dplyr::count(chromosome, window_start, window_end, name = "gene_count")


# Correct averaging formula
pi_combined <- pi_combined %>%
  group_by(chromosome, window_start, window_end) %>%
  dplyr::summarise(avg_pi_4_fold_window = sum((count_diffs_4fold)) / sum(count_comparisons_4fold),
            avg_pi_0_fold_window = sum((count_diffs_0fold)) / sum(count_comparisons_0fold))


# Merge gene density back into pi_combined
pi_combined <- left_join(pi_combined, gene_density, by = c("chromosome", "window_start", "window_end"))

head(pi_combined)
```
#Upload selection coefficient optimization

```{r}
chr1_one_coeff <- read.table("/groups/nordborg/projects/genome_organization/03_processing/008_modeling_selection/Chr3_two_coeff_0.0005251216490145434_0.00032713778498542727_data_with_te.csv", sep=",", header = T)
```
# Visualization
```{r}

library(dplyr)
library(ggplot2)
library(scales)

minimal_theme_settings <- ggplot2::theme_minimal(base_size = 25)
custom_theme <- theme(
  plot.background = element_blank(), 
  strip.text = element_text(size = 25, color = "gray30"),
  panel.background = element_blank(),
  panel.border = element_blank(), # Ensure no frame
  panel.grid = element_blank(),
  axis.title.x = element_text(color = "gray30", size = 25), # Set x-axis title color and size
  axis.title.y = element_text(color = "gray30", size = 25), # Set y-axis title color and size
  axis.ticks.x = element_blank(),
  axis.line = element_line(color = "gray30"), # Add this line to show x and y axis lines
  strip.background = element_blank(),
   axis.text.x = element_text(color = "gray30", size = 25), # Adjust x-axis text size and color
    axis.text.y = element_text(color = "gray30", size = 25),
  strip.text.x = element_text(color = "gray30",size = 25), # Control color of facet wrap titles
  legend.text = element_text(color = "gray30",size = 25) # Set legend text color
)

# Assuming your data frames are correctly named and loaded

# Since the structure might be slightly different, let's assume we've already
# aligned `mean_rho_per_window` with `product_ET` and `pi_combined` by `window_start`
# This step might require adjustments based on your real data structure


data_combined <- chr1_one_coeff %>%
  select(chromosome, window_start, window_end, product_ET, avg_pi) %>%
  inner_join(pi_combined, by = c("chromosome", "window_start", "window_end")) %>%
  # Correct the join condition here
  inner_join(mean_rho_per_window, by = c("chromosome", "window_start" = "Start"))


# Scaling the data for pi, ET (product_ET), and rho (Mean_rho_mean)
data_scaled <- data_combined %>%
  mutate_at(vars(avg_pi, product_ET, Mean_rho_mean), scale)

# Renaming for clarity in plotting
data_long <- data_scaled %>%
  select(window_start, avg_pi_scaled = avg_pi, ET_scaled = product_ET, rho_scaled = Mean_rho_mean) %>%
  pivot_longer(-window_start, names_to = "Variable", values_to = "Scaled Value")

data_long_1 <- data_long[data_long$Variable!="rho_scaled",]
data_long_2 <- data_long[data_long$Variable!="ET_scaled",]
# Plotting with ggplot2
library(scales)


plot1 <- ggplot(data_long_1, aes(x = window_start, y = `Scaled Value`, color = Variable)) +
  geom_rect(data = data_long_2, aes(xmin = 17559778 + 4000000, xmax = 26408303, ymin = -Inf, ymax = Inf), 
            fill = "#DDDDDD", alpha = 0.03, color = "grey") +  # Set border color to grey
  geom_rect(aes(xmin = ifelse(14841109 - 4000000 < 1, 1, 14841109 - 4000000), xmax = 17559778 + 4000000, ymin = -Inf, ymax = Inf), 
            fill = "#DDDDDD", alpha = 0.03, color = "grey") +  # Set border color to grey
  geom_vline(data = data_long_2, aes(xintercept = 26408303), color = "gray30", linetype = "solid", size = 0.8, alpha = 0.3) +
  geom_vline(aes(xintercept = ifelse(14841109 - 4000000 < 1, 1, 14841109 - 4000000)), color = "gray30", linetype = "solid", size = 0.8, alpha = 0.3) +
  geom_line(size = 1, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.7) +
  scale_x_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6)) +  # Consolidated scale_x_continuous
  labs(x = "Window Start", y = "Scaled Value", title = "Scaled Pi, ET, and Rho across Chromosome 1 Windows") +
  theme(legend.title = element_blank()) +
  scale_color_manual(values = c("avg_pi_scaled" = "#51829B", "ET_scaled" = "#FF6C22")) +
  custom_theme  # Ensure custom_theme is correctly defined earlier in your code

width_inches <- 25
height_inches <- 8
dpi <- 600
CairoPNG("/groups/nordborg/projects/genome_organization/08_manuscripts/01_figures/et_vs_rho.png", width = width_inches * dpi, height = height_inches * dpi, dpi = dpi)
print(plot1)
dev.off()

```







